---
title: "Anatomy of the model: the generative process"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anatomy of the model: the generative process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5, dpi = 96)
library(bayespulse)
```

An observed hormone series is the sum of a baseline level, pulsatile secretion
(a set of Gaussian pulses), exponential elimination back toward baseline, and
measurement error. This vignette builds intuition for that generative process
by simulating series and varying each piece in turn. The formal model, with its
secretion and elimination functions written out, is given in the getting-started
vignette; here the goal is simply to see what each parameter does to the shape
of the series.

Every simulation below uses `simulate_pulse()`'s default
`pulse_distribution = "truncnorm"`, which puts pulse mass and width on their
natural scale -- a `mass_mean` of 3.5 really is a mean mass of 3.5. That keeps
the sweeps easy to read. When you go on to *fit* a model, `pulse_spec()` defaults
to the lognormal parameterization instead, where the same quantities live on the
log scale; the getting-started and priors vignettes cover that difference.

## A single simulated series

We start from one realistic series to anchor the comparisons that follow.

```{r base-series}
set.seed(1)
sim <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = 3.5,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = 45)
plot(sim)
```

## Pulse mass

Pulse mass controls how much hormone is released per pulse: a larger mass
produces taller pulses.

```{r sweep-mass, fig.height = 7}
masses <- c(1.5, 3.5, 6.0)
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (m in masses) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = m,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = 45)
  plot(s$data$time, s$data$concentration, type = "l",
       main = paste("mass_mean =", m), xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

## Pulse width

Pulse width controls how spread out each secretion event is: a larger width
gives broader, flatter humps.

One detail matters enormously when reading these plots: **width is on the
variance scale**. A `width_mean` of 100 is a pulse standard deviation of
$\sqrt{100} = 10$ minutes, not 100 minutes. That square root compresses things
sharply, and it is easy to sweep a range that looks like it should matter and
see almost no change. Sampling here is every 10 minutes, so any pulse whose
standard deviation is well under 10 minutes is narrower than the sampling grid
can resolve -- widths of 15 and 60 (SD 3.9 and 7.7 minutes) produce series that
are nearly indistinguishable. To see the effect we need a range that straddles
the sampling interval:

```{r sweep-width, fig.height = 7}
widths <- c(4, 100, 625)   # variance scale: SD = 2, 10, and 25 minutes
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (w in widths) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = 3.5,
                      width_mean = w, constant_baseline = 2.6,
                      constant_halflife = 45)
  plot(s$data$time, s$data$concentration, type = "l",
       main = sprintf("width_mean = %d  (SD = %.0f min)", w, sqrt(w)),
       xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

Now the progression is clear: at SD 2 minutes each pulse is a sharp spike that
the sampling grid barely catches, while at SD 25 minutes secretion is smeared
across several samples and neighbouring pulses begin to merge into a smooth,
low-amplitude wave. This is a real limit on what any deconvolution model can
recover -- pulses much narrower than the sampling interval leave too little
trace in the data to be identified reliably, a point the study-design vignette
returns to.

## Half-life

The elimination half-life controls how quickly hormone clears between pulses:
a longer half-life means slower decay and higher troughs.

```{r sweep-halflife, fig.height = 7}
halflives <- c(20, 45, 90)
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (h in halflives) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = 3.5,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = h)
  plot(s$data$time, s$data$concentration, type = "l",
       main = paste("constant_halflife =", h), xlab = "Time (min)",
       ylab = "Conc.")
}
par(op)
```

## Inter-pulse interval

The mean inter-pulse interval controls how often pulses occur: a larger
interval yields fewer, more widely separated pulses.

```{r sweep-ipi, fig.height = 7}
ipis <- c(8, 14, 24)
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (ip in ipis) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, ipi_mean = ip,
                      ipi_var = 10, ipi_min = 4, mass_mean = 3.5,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = 45)
  plot(s$data$time, s$data$concentration, type = "l",
       main = paste("ipi_mean =", ip), xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

Together these four parameters, plus the baseline level and the measurement
error variance, generate the rich variety of shapes seen in real hormone
series. The next vignettes show how to encode prior knowledge about each of
them and how to fit the model to recover them from data.
