---
title: "Specifying priors and starting values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying priors and starting values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5, dpi = 96)
library(bayespulse)
```

In a Bayesian deconvolution model the priors encode what we know about the
hormone system *before* seeing the data. The observed series is decomposed into
a baseline, a set of pulses (each with a mass and a width), exponential
elimination, and measurement error. Many of these pieces are only weakly
identified from a single subject's data -- a tall narrow pulse and a shorter
wider one can produce similar-looking peaks, for example -- so reasonably
informative priors are not just a formality; they stabilize estimation and keep
the sampler in physiologically plausible regions.

`pulse_spec()` is where all of this is set: the priors themselves, the starting
values that seed the MCMC chains (the `sv_*` arguments), and the proposal
variances that tune Metropolis-Hastings steps (the `pv_*` arguments). This
vignette focuses on the priors and what they mean physiologically; starting
values and proposal-variance tuning are covered by the fitting and diagnostics
material. Because nothing here runs MCMC -- we only draw from priors and
simulate -- the examples are fast.

## First: which parameterization?

Before reading any single prior, you need to know which scale the model is on,
because it changes what every mass and width number *means*. Three arguments set
this, and their defaults are the parameterization used in the source papers:

| Argument | Default | Effect |
|---|---|---|
| `pulse_distribution` | `"lognormal"` | Pulse mass and width random effects are Gaussian on the **log** scale. `"truncnorm"` restores the legacy natural-scale truncated normal. |
| `sd_prior` | `"uniform"` | Pulse-to-pulse SDs get a Uniform(0, max) prior, so `prior_sd_mass` is an **upper bound**. Under `"half_cauchy"` the same argument is a **scale** instead. |
| `student_t_pulses` | `FALSE` | Pulse random effects are Gaussian. `TRUE` adds a per-pulse t-scale mixture, giving heavier tails. |

The choice also selects which set of numeric defaults fills in any prior,
starting value, or proposal variance you leave unspecified -- log-scale defaults
under `"lognormal"`, natural-scale ones under `"truncnorm"`. Mixing scales is the
easiest mistake to make here: passing `prior_mass_mean = 2.5` to a lognormal
model does not request a mean mass of 2.5, it requests a mean *log* mass of 2.5,
i.e. a median mass of $e^{2.5} \approx 12$.

To reproduce the package's pre-lognormal behaviour exactly, set all three
together: `pulse_spec(pulse_distribution = "truncnorm", sd_prior = "half_cauchy",
student_t_pulses = TRUE)`.

## What the priors mean

Each prior controls one part of the generative process. Under the default
lognormal parameterization the mass and width arguments are on the **log** scale;
baseline, half-life, and pulse count are always on their natural scale. Variances
express how tightly a quantity is constrained a priori.

| Argument | Controls | Scale | Typical choice / notes |
|---|---|---|---|
| `prior_baseline_mean`, `prior_baseline_var` | Constant baseline concentration | natural | Center near the assay's observed trough; a moderate variance lets the data adjust it. |
| `prior_mass_mean`, `prior_mass_var` | Mean pulse mass (hormone released per pulse) | log | Set from prior studies of the hormone. A value of 1.2 is a median mass of about 3.3. |
| `prior_width_mean`, `prior_width_var` | Mean pulse width (spread of a pulse) | log of a **variance** | Doubly indirect: 3.5 is a variance of $e^{3.5} \approx 33$, i.e. a pulse SD of about 5.8 minutes. |
| `prior_sd_mass` | Pulse-to-pulse variability in mass | log | Under the default `sd_prior = "uniform"` this is the upper bound of the prior, not a standard deviation. Larger values admit more heterogeneity. |
| `prior_sd_width` | Pulse-to-pulse variability in width | log | Estimated from the data like any other parameter. Interpreted the same way as `prior_sd_mass`. |
| `prior_halflife_mean`, `prior_halflife_var` | Elimination half-life | natural (minutes) | Often known fairly well from physiology, so a smaller variance is reasonable. |
| `prior_mean_pulse_count` | Expected number of pulses over the window | natural (count) | Anchors how many secretion events to expect; scale to the observation length. |
| `location_prior_type`, `prior_location_gamma`, `prior_location_range` | Spatial prior on pulse locations (e.g. `"strauss"`) | natural (minutes) | Repulsive prior; see "The location prior" below. |

Two of these deserve a second look. **Pulse width is on the log of a variance**,
so it is two transformations away from the "how many minutes long is a pulse"
question you probably want to ask; the anatomy vignette shows how easily that
trips up a parameter sweep. And **`prior_sd_mass` / `prior_sd_width` are not
standard deviations** under the default uniform prior -- they are the upper end
of the range the SD is allowed to take.

Unless you have a specific reason to override them, leaving the mass and width
arguments unset is usually the better choice: the built-in log-scale defaults are
tuned for this parameterization, and the individual-pulse proposal variances in
particular were calibrated by simulation study. Hand-picked values carried over
from the natural-scale parameterization tend to mix poorly.

## Informative vs vague priors

The variance arguments are the main dial between informative and vague. A small
prior variance encodes confidence -- the model is told a parameter is near a
particular value and the data have to work hard to move it (informative). A
large variance expresses ignorance and lets the data dominate (vague). The
right choice depends on how much trustworthy prior knowledge you have: half-life
is often well characterized and warrants a tighter prior, while baseline or
pulse mass may deserve more room.

A good way to see what a prior *implies* is a prior-predictive check: draw
parameters from the prior, simulate series, and look at the spread of outcomes.
Below we vary the baseline and half-life priors only, drawing each from a normal
and flooring at small positive values so `simulate_pulse()` always receives
valid inputs. Both are natural-scale parameters, so this example sidesteps the
log-scale question entirely and we can leave `simulate_pulse()` on its default
natural-scale settings.

```{r prior-predictive, fig.height = 6}
set.seed(2026)
draw_series <- function(baseline_sd, halflife_sd) {
  bl <- max(0.3, rnorm(1, mean = 2.5, sd = baseline_sd))
  hl <- max(5,   rnorm(1, mean = 45,  sd = halflife_sd))
  simulate_pulse(num_obs = 60, interval = 10, mass_mean = 2.5,
                 width_mean = 35, constant_baseline = bl,
                 constant_halflife = hl)
}
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
# Informative priors: small SDs -> series cluster tightly
for (i in 1:2) {
  s <- draw_series(baseline_sd = 0.25, halflife_sd = 4)
  plot(s$data$time, s$data$concentration, type = "l",
       main = "informative prior", xlab = "Time (min)", ylab = "Conc.")
}
# Vague priors: large SDs -> series vary widely
for (i in 1:2) {
  s <- draw_series(baseline_sd = 1.2, halflife_sd = 25)
  plot(s$data$time, s$data$concentration, type = "l",
       main = "vague prior", xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

The two informative-prior draws (top row) look much alike: the prior tightly
constrains where the series can sit and how quickly it decays. The two
vague-prior draws (bottom row) range widely, because the prior offers little
constraint and very different baselines and decay rates are equally admissible.
Inspecting these prior-predictive series is a quick sanity check that your
priors imply plausible data before any fitting begins. This example varies only
the baseline and half-life for illustration, but the same logic applies to the
mass, width, and pulse-count priors.

## The location prior

Pulse locations get a spatial prior rather than independent draws. The default
`location_prior_type = "strauss"` uses a Strauss point process, which is
*repulsive*: it down-weights configurations in which pulses sit very close
together. This matches the biology -- real secretion events do not pile
arbitrarily close to one another, since the system needs time between bursts.
Two arguments tune it: `prior_location_gamma` is the repulsion strength (a value
of $\gamma = 1$ means no repulsion, so locations are effectively
order-independent, while smaller $\gamma$ imposes stronger repulsion that
discourages near-coincident pulses), and `prior_location_range` is the
interaction distance, the separation below which pulses are penalized.

## Where to go next

`population_spec()` mirrors these same prior ideas at the population level,
where the priors describe distributions of pulse parameters across subjects. The
getting-started vignette shows a complete `pulse_spec()` call in a fitted model,
and the diagnostics material covers starting values and convergence in detail.

Priors matter most where the data are least able to speak, and for this model
that is the mass/width ridge described at the top of this vignette. Once you have
a fitted model, `identifiability_check()` reports how much of each parameter's
posterior is being carried by its prior rather than the data -- a direct way to
tell whether a prior you intended as mildly informative is in fact doing the
bulk of the work. The convergence vignette puts it to use.
